The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-09-24 conf_article/neon.tex
did some editing, mainly in introduction

In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the Ky Fan k-norms to introduce a new family of Muon-like algorithms we name {\it Fanions}. Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.

@@

\section{Introduction}
This is the original text. \added[id=IK, comment={I think we should mention this.}]{This is a new sentence that has been added.}

@@

 That is what can be said from a practical point of view. From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update \citep{bernstein2025deriving}. Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

 In this article, we focus on the two most common matrix norms akin to the spectral norm, namely the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.

@@

\section{Matrix side of updates}
To compute the algorithms' updates, we use thick-restarted Lanczos method for singular value problem (TRLan) on $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ matrices (the one with less size is picked), implemented in CuPy lSibrary \citep{cupy_svds_ref} and described in \citet{simonz1998thick}.

@@

    The most pathological case is $\alpha < 0$, which corresponds to the lmo with an area that is not a norm ball! Despite this violation, the mixture of algorithms reaches almost the same accuracy as vanilla Muon.

>>>>

In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a \deleted[id=IK, comment={Redundant}]{new} family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. \replaced[id=IK, comment={This rewrite clarifies the sentence structure, making it clear that the norms (not the algorithms) are being combined.}]{Subsequently, we construct a second family of algorithms, dubbed {\it F-Fanions}, based on the duals of convex combinations of Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muon}.}{Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}.} We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, \replaced[id=IK, comment={Alex, clarify this please. It's a very non-obvious and non-clear conclusion, especially when knowing that Muon itself only provides an approximate answer to the lmo.}]{which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.}

@@

\section{Introduction}

@@

That is what can be said from a practical point of view. \replaced[id=IK, comment={Rewrote this to be more precise. The innovation wasn't just 'using' a norm, but deriving the update from a principled, constrained optimization problem. I think this better reflects the contribution of the Muon paper.}]{From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerges as the solution to an optimization problem constrained by the spectral norm \citep{bernstein2025deriving}}{From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update}. Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

In this article, \replaced[id=IK, comment={I think this is much clearer to introduce all algorithms like that, the paragraphs where one of every two words is a ***on are confusing.}]{we generalize the spectral-norm-based approach by first leveraging the family of Ky Fan k-norms to derive a new class of algorithms, which we call Fanions. Subsequently, we create a second, hybrid family named F-Fanions, by constructing convex combinations of the Ky Fan norms with the Frobenius norm and taking dual of that composite norm. Working within the linear minimization oracle (lmo) framework~--- which is equivalent (up to a factor) to the trust region approach~--- we derive the explicit update formulas for both algorithm families. Later in \cref{sec:matrix-side} we explain how their updates can be approximated with the same $\cO(m n \min\{m,n\})$ complexity as Muon updates are computed via Newton-Schulz iterations.}{we focus on the two most common matrix norms akin to the spectral norm, namely the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.}

@@

\section{Matrix side of updates}\label{sec:matrix-side}
To compute the algorithms' updates, we use thick-restarted Lanczos method for singular value problem (TRLan) on $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ matrices (the one with less size is picked), implemented in CuPy lSibrary \citep{cupy_svds_ref} and described in \citet{simonz1998thick}.

@@

    The most pathological case is \replaced[id=IK, comment={In the pictures there are no experiments with $\alpha < 0$, but there is a one with $\alpha $}]{$\alpha > 1$}{$\alpha < 0$}, which corresponds to the lmo with an area that is not a norm ball! Despite this violation, the mixture of algorithms reaches almost the same accuracy as vanilla Muon.
